Constraint Grammar as a SAT problem

### 2015-05-31 paper/nodalida2015.tex
more and less and better and worse

This behaviour is explained by simple properties of logical formulae.
When the rules are applied to the text, they are translated into
implications: \texttt{REMOVE prn IF (1 n)} becomes
 \emph{casa}\texttt{<n>} $\Rightarrow$ $\neg$\emph{la}\texttt{<prn>},
 which reads ``if the \texttt{n} reading for \emph{casa} is true, then
 discard the \texttt{prn} reading for \emph{la}''.
Any implication $a\,\Rightarrow\,b$ can be represented as a disjunction
$\neg a\,\vee\,b$; intuitively, either the antecedent is false
and the consequent can be anything, or the consequent is true and the
antecedent can be anything.
Due to this property, our rule translates into the disjunction 
$\neg$\emph{casa}\texttt{<n>} $\vee$ $\neg$\emph{la}\texttt{<prn>}.

However, our rule is not the only way to reach this form: since
disjunction doesn't care about order, we notice that also
\emph{la}\texttt{<prn>} $\Rightarrow$ $\neg$\emph{casa}\texttt{<n>} 
corresponds to the same disjunction.
This means that the rules are logically flipped: \texttt{REMOVE prn IF
  (1 n)} translates into the same logical formula as  \texttt{REMOVE n
  IF (-1 prn)}. 
A rule with more conditions corresponds to many
rules, each condition taking its turn to be the target of removal or
selection.

@@

The traditional CG relies on that applying the rule set iteratively:
some rules fire during the first iteration, either because their
conditions don't require cautious context, or because some words are
unambiguous to start with. This makes some more words unambiguous, and
new rules can fire during the second iteration.

In SAT-CG, the notion of cautious context is irrelevant.
When the rules are applied to the text, they are translated into
implications, such as \\ 
 \emph{casa}\texttt{<n>} $\Rightarrow$ $\neg$\emph{la}\texttt{<prn>},
 which reads ``if the \texttt{n} reading for \emph{casa} is true, then
 discard the \texttt{prn} reading for \emph{la}''. Each rule generates
 a number of these implications, and the SAT solver tries to find a model that will satisfy them.

@@

Contrast this with the behaviour of the standard CG.

With the new rule, standard CG will also remove the verb reading from \emph{casa}, 
but it is in no way connected to the choice for \emph{la}. It all
depends of the order of the two rules; if the \texttt{det} reading of \emph{la} is removed first, then we are stuck with that choice.

@@

\begin{itemize}
\item [\texttt{1)}] \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{REMOVE verb IF (-1 adj) ;} \\
         \texttt{SELECT verb ;}
\item [\texttt{2)}] \texttt{SELECT verb ;} \\
         \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{REMOVE verb IF (-1 adj) ;} 
\end{itemize}

Without order, both of these rule sets will conflict, if applied to an
input that has sequence \texttt{det verb} or \texttt{adj verb}.
The SAT solver is fed clauses that tell to select a verb and remove a
verb, and it cannot build a model that satisfies all of those clauses.
To solve this problem, we create a variable for every clause
(\emph{instance of rule application}), and request a solution where maximally many of these variables are true.
If there is no conflict, then the maximal solution is one where all of
these variables are true; that is, all instances of rule application are kept.
In case of a conflict, SAT solver makes it possible to discard only
minimal amount of rule applications. For the example, assuming that
there are instances for both contexts for the REMOVE rules, then just
ignoring the SELECT rule will lead to maximal solution. 
This corresponds loosely to the common design pattern in
CGs, where there is a number of rules with the same target, ordered
such that more secure rules come first,
with a catch-all rule with no condition as the last resort, to be
applied if none of the previous has fired.

@@

The design of unordered rules can lead to new ways of writing CGs,
potentially with less mental effort to the writer, but also less
transparent. By testing against gold standard, we see that the
unordered scheme not optimal for applying ready-made
CGs, where the order is a crucial detail.
We can also emulate order: enter clauses produced by each rule one by
one, and assume the solver state reached so far is correct; if a
conflict is introduced by new clauses, discard them and move on to
next.
This is also better for the performance, because the maximisation
problems are smaller.

@@

The chosen heuristic produced the best results with examples in the order of tens or hundreds of rules.
Larger rule sets are yet to be tested.

@@

When conducting our tests, we have been able to construct a small grammar 
which, when run with the unordered scheme, disambiguates more than the
same grammar run with VISL CG-3. However, when tested with grammars
that are written for the original CG, SAT-CG loses with both ordering
strategies, but emulating order fares better. 
This suggests that our implementation should not compete with existing
state-of-the-art, but rather it has value as a way of relating the CG
formalism to wider context in the theory of computer science.
In Section~\ref{sec:apps} we discuss more about possible applications.

\section{Results and evaluation}
\label{sec:eval}
\paragraph{Time}

@@

Table~\ref{table:time} shows the execution time compared to VISL CG-3. 
The number of rules in the grammar is more significant than the word
count; this holds also for the performance of VISL CG-3.
From the SAT solving side, maximisation is the most costly operation;
hence emulating ordering is faster, because maximisation problems are smaller.
Parsing Don Quijote (384,155 words) with the rule set of 261 rules,
the function was called 147,253 times, and with 19 rules, 132,255 times.
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
The difference in the execution times suggests that there are other
reasons for the worse performance---this is to be expected,
SATCG is currently just a naive proof-of-concept implementation with no optimisations.

\begin{table}
  \centering
  \begin{tabular}{|l|c|c|c|c|}
     \hline
   \textbf{\# words} & \textbf{\# rules} &  \textbf{SATCG\textsubscript{u}} & \textbf{SATCG\textsubscript{o}} & \textbf{VISL CG-3} \\ \hline
                384,155  & 19   & TODO &  23.1s   & 4.2s\\ %\hline
	        384,155  & 99   & TODO & 1m40.6s & 6.1s \\ %\hline
                384,155  & 261  & TODO & 2m36.7s & 10.7s \\ \hline
  \end{tabular}
  \caption{Execution times.}
  \label{table:time}
\end{table}

\paragraph{Performance against VISL CG-3}

We took a tagged corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged} of 21865 words for Spanish, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We only took select and remove rules (no substitute/iff), after which the constraint grammar had 261 rules.
With this setup, we took the text of the tagged corpus with all
ambiguities, and ran both SATCG and VISL CG-3 with the same grammar. 
Treating the corpus as the gold standard, the disambiguation by
SATCG achieves 75.09 \% correct result and VISL CG-3 75.80 \%.
Similar patterns were observed with small (\textless{}20) rule sets
written by the authors; depending on the subset, SATCG and VISL CG-3 had
a difference of at most $\Mypm$ 1.5 \%. 
Introducing rules one by one up to 19, the
performance improved in a very similar rate, with less than 0.5 \%
difference between the systems at each new rule.

% Additional tests could include plugging SATCG into Apertium
% translation pipeline, and comparing the translation quality.

\section{Applications and Future Work}
\label{sec:apps}
As described in Section~\ref{sec:eval}, our implementation is not
competetive with the state of the art. It is more fruitful to focus on
things where the SAT-based representation can make an impact.
One such thing is grammar analysis.

Traditionally, CG rules are applied deterministically and
sequentially; at the execution of rule, there is no way to
go back to earlier rules and undo them. Thus there is no way to know
what part of the input comes from the morphological analyser and what
part from applying previous rules.

@@

The SAT-CG implementation is ideal for this kind of corpus-based grammar
analysis. After applying each rule, it stores the clauses produced by
it and commits to them -- this makes the execution slow, but it
retains the dependencies.
Given the design that each rule application (instead of rule) gets a variable,
in case of a conflict, the program prints out the particular
application that violates the previous clauses, with the sentence
where it is applied. 
This conflict solving can be done per section, or for the whole grammar. 
In case of section-based, we run each section on the input, and give
the output as an input to the new section, discarding all clauses.

@@

For future work, we intend to apply the SAT solver in testing for
conflicts without a corpus. Let us illustrate the idea with the rule
set shown in Section 3:

\begin{itemize}
\item [\texttt{1)}] \texttt{SELECT verb ;} \\
         \texttt{REMOVE verb IF (-1 det) ;} \\

\end{itemize}

\begin{itemize}
\item [\texttt{2)}] \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{SELECT verb ;}
\end{itemize}

Assume we have tag set \texttt{\{det, noun, verb, sg, pl\}}. Then we
can ask the SAT solver a question if there exists an input which
satisfies our constraints: both rules, applied in the order, remove
something from the input.
after applying the first rule, can be modified by the second rule.

Implementing the design on a larger scale is left for future work.

@@

SAT-solvers are nowadays powerful enough to be used for dealing with Constraint Grammar. A logic-based approach to CG has possible advantages over more traditional approaches; a SAT-solver may disambiguate more words, and may do so more precisely. Also, the SAT-solver requires less rules, and these rules are simpler. The ordering of rules is something we found was incompatible with a logic-based approach. We compensate this by maximising rule applications. Whether this is an advantage or disadvantage remains to be seen. Our initial experimental results are promising.

@@

We thank Tino Didriksen, Eckhard Bick, Francis Tyers and Anssi
Yli-Jyrä for comments and suggestions; also everyone at the CG
workshop who participated in the discussion.

>>>>

This behaviour is explained by simple properties of logical formulae.
When the rules are applied to the text, they are translated into
implications: \texttt{REMOVE prn IF (1 n)} becomes
 \emph{casa}\texttt{<n>} $\Rightarrow$ $\neg$\emph{la}\texttt{<prn>},
 which reads ``if the \texttt{n} reading for \emph{casa} is true, then
 discard the \texttt{prn} reading for \emph{la}''.
Any implication $a\,\Rightarrow\,b$ can be represented as a disjunction
$\neg a\,\vee\,b$; intuitively, either the antecedent is false
and the consequent can be anything, or the consequent is true and the
antecedent can be anything.
Due to this property, our rule translates into the disjunction 
$\neg$\emph{casa}\texttt{<n>} $\vee$ $\neg$\emph{la}\texttt{<prn>}.
%Since disjunction doesn't care about the order, we notice that
However, our rule is not the only way to reach this form---the implication
\emph{la}\texttt{<prn>} $\Rightarrow$ $\neg$\emph{casa}\texttt{<n>} 
corresponds to the same disjunction.
This means that the rules are logically flipped: \texttt{REMOVE prn IF
  (1 n)} translates into the same logical formula as  \texttt{REMOVE n
  IF (-1 prn)}. 
A rule with more conditions corresponds to many
rules, each condition taking its turn to be the target of removal or
selection.

@@

Traditional CG relies on that applying the rule set iteratively:
some rules fire during the first iteration, either because their
conditions don't require cautious context, or because some words are
unambiguous to start with. This makes some more words unambiguous, and
new rules can fire during the second iteration.

In SAT-CG, the notion of cautious context is irrelevant. Instead of
removing readings immediately,  each rule generates a  number of
implications, and the SAT solver tries to find a model that will satisfy them.

@@

Contrast this with the behaviour of the standard CG.
With the new rule, standard CG will also remove the verb reading from \emph{casa}, 
but it is in no way connected to the choice for \emph{la}. It all
depends of the order of the two rules; if the \texttt{det} reading of
\emph{la} is removed first, then we are stuck with that choice.
The SAT-based approach only removes readings after it has enough
evidence to do that. From the grammar writer perspective, this removes
a burden of having to decide whether the rule should be cautious or
not---the SAT solver will only apply it if it has enough evidence from
the surrounding context.

@@

\begin{itemize}
\item [\texttt{1)}] \texttt{REMOVE v IF (-1 det) ;} \\
         \texttt{REMOVE v IF (-1 adj) ;} \\
         \texttt{SELECT v ;}
\item [\texttt{2)}] \texttt{SELECT v ;} \\
         \texttt{REMOVE v IF (-1 det) ;} \\
         \texttt{REMOVE v IF (-1 adj) ;} 
\end{itemize}

Without order, both of these rule sets will conflict, if applied to an
input that has sequence \texttt{det v} or \texttt{adj v}.
The SAT solver is fed clauses that tell to select a verb and remove a
verb, and it cannot build a model that satisfies all of those clauses.
To solve this problem, we create a variable for every clause
(\emph{instance of rule application}), and request a solution where maximally many of these variables are true.
If there is no conflict, then the maximal solution is one where all of
these variables are true; that is, all instances of rule application
are kept.

In case of a conflict, SAT solver makes it possible to discard only
minimal amount of rule applications. Continuing with the example,
assume that there are instances for both contexts for the REMOVE
rules, then just ignoring the SELECT rule will lead to maximal solution. 
This corresponds loosely to the common design pattern in
CGs, where there is a number of rules with the same target, ordered
such that more secure rules come first,
with a catch-all rule with no condition as the last resort, to be
applied if none of the previous has fired.

@@

The design of unordered rules can lead to new ways of writing CGs,
potentially with less mental effort to the writer, but also less
transparent. By testing against gold standard, we see that the
unordered scheme not optimal for applying ready-made
CGs, where the order is a crucial detail.
We can also emulate order: enter clauses produced by each rule one by
one, and assume the solver state reached so far is correct; if a
conflict is introduced by new clauses, discard them and move on to
next.
This is also better for the performance, because the maximisation
problems are smaller. \\

@@

@@

When conducting our tests, we have been able to construct a small grammar 
which, when run with the unordered scheme, disambiguates more
accurately than the same grammar run with VISL CG-3. However, when
testing with grammars that are written for the original CG, SAT-CG
loses with both ordering strategies, but emulating order fares better.
This suggests that our implementation should not compete with existing
state-of-the-art, but rather it has value as a way of relating the CG
formalism to wider context in the theory of computer science.
In Section~\ref{sec:apps} we discuss more about possible applications.

\section{Results and evaluation}
\label{sec:eval}

\subsection{Time}

@@

Table~\ref{table:time} shows the execution time compared to VISL CG-3. 
The number of rules in the grammar is more significant than the word
count; this holds also for the performance of VISL CG-3.
From the SAT solving side, maximisation is the most costly operation; emulating ordering is faster, because maximisation problems are smaller.
Parsing Don Quijote (384,155 words) with the rule set of 261 rules,
the function was called 147,253 times, and with 19 rules, 132,255 times.
%However, with both rule sets, half of the sentences in Don Quijote
%needed less than 6 calls of maximise, and 75 \% of the sentences
%needed less than 14.
The difference in the execution times suggests that there are other
reasons for the worse performance---this is to be expected,
SAT-CG is currently just a naive proof-of-concept implementation with no optimisations.

\begin{table}
  \centering
  \begin{tabular}{|c|c|c|c|}
     \hline
     \textbf{\# rules} &  \textbf{SAT-CG\textsubscript{u}} & \textbf{SAT-CG\textsubscript{o}} & \textbf{VISL CG-3} \\ \hline
      19   & TODO &  23.1s   & 4.2s\\ %\hline
      99   & TODO & 1m40.6s & 6.1s \\ %\hline
      261  & TODO & 2m36.7s & 10.7s \\ \hline
  \end{tabular}
  \caption{Execution times.}
  \label{table:time}
\end{table}

\subsection{Performance against VISL CG-3}

We took a tagged corpus\footnote{https://svn.code.sf.net/p/apertium/svn/branches/apertium-swpost/apertium-en-es/es-tagger-data/es.tagged} of 21865 words for Spanish, 
and a small constraint grammar\footnote{https://svn.code.sf.net/p/apertium/svn/languages/apertium-spa/apertium-spa.spa.rlx}, produced independently of the authors.
% We ignored substitute rules, but kept subreadings and unification, even though our implementation doesn't handle them.
We only took select and remove rules (no substitute/iff), after which the constraint grammar had 261 rules.
With this setup, we took the text of the tagged corpus with all
ambiguities, and ran both SAT-CG and VISL CG-3 with the same grammar. 
Treating the corpus as the gold standard, the disambiguation by
SAT-CG achieves F-score of 81.54 \%  and VISL CG-3 82.56 \%.
Similar patterns were observed with small (\textless{}20) rule sets
written by the authors; depending on the subset, SAT-CG and VISL CG-3 had
a difference of at most $\Mypm$ 1.5 \%. 
Introducing rules one by one up to 19, the
performance improved in a very similar rate, with less than 0.5 \%
difference between the systems at each new rule.

% Additional tests could include plugging SAT-CG into Apertium
% translation pipeline, and comparing the translation quality.

\section{Applications and Future Work}
\label{sec:apps}
As described in Section~\ref{sec:eval}, our implementation is not
competetive with the state of the art. It is more fruitful to focus on
things where the SAT-based representation can make an impact.
One such thing is grammar analysis.
There has been work on automatic tuning of hand-written CGs \todo{cite
Eckhard}, but no tools which would search for inconsistencies or
unoptimal design in order to point it out to the grammar writer.

Traditional CG rules are applied deterministically and
sequentially; at the execution of rule, there is no way to
go back to earlier rules and undo them. Thus there is no way to know
what part of the input comes from the morphological analyser and what
part from applying previous rules.

@@

SAT-CG works for this kind of corpus-based grammar analysis as a side
effect of its design. After applying each rule, it stores the clauses
produced by it and commits to them---this meas that it retains the
dependencies.
Given the design that each rule application (instead of rule) gets a variable,
in case of a conflict, the program prints out the particular
application that violates the previous clauses, with the sentence
where it is applied. 
This conflict solving can be done per section, or for the whole grammar. 
In case of section-based application, we run each section on the
input, and give the output as an input to the new section, discarding
all clauses. Alternatively, we can let the rules interact with each
other regardless of the section boundaries.

@@

For future work, we intend to apply the SAT solver for corpus-free
conflict testing. Let us illustrate the idea with the rule
set shown in Section 3:

\begin{itemize}
\item [\texttt{1)}] \texttt{REMOVE verb IF (-1 det) ;} \\
         \texttt{SELECT verb ;}

\item [\texttt{2)}] \texttt{SELECT verb ;} \\
         \texttt{REMOVE verb IF (-1 det) ;} 
\end{itemize}

Assume we have tag set \texttt{\{det, noun, verb, sg, pl\}}. Then we
can ask the SAT solver a question if there exists an input which
satisfies our constraints: both rules, applied in the order, remove
something from the input.
SAT solver tells us that with sentence length 2 and maximum 4 readings
for the whole sentence (1+3, 2+2 or 3+1), there are several inputs that
satisfy the constraints for the first rule set, but none that
satisfies the second rule set. Thus we can say that the first rule set
is sane, but the second rule set is conflicting.

Implementing and testing this on a larger scale is left for future work.

@@

SAT-solvers are nowadays powerful enough to be used for dealing with
Constraint Grammar. A logic-based approach to CG has possible
advantages over more traditional approaches; a SAT-solver may
disambiguate more words, and may do so more precisely. Also, the
SAT-solver requires less rules, and these rules are simpler. 
\todo{proper conclusions for the new stuff}
% The ordering of rules is something we found was incompatible with a logic-based approach. We compensate this by maximising rule applications. Whether this is an advantage or disadvantage remains to be seen. Our initial experimental results are promising.

@@

We thank Tino Didriksen, Eckhard Bick, Francis Tyers and Anssi
Yli-Jyr{\"a} for comments and suggestions, samt the anonymous
reviewers and everyone who participated in the discussion at the CG workshop.
